% COM3013 Coursework Option A Report
% Left-aligned sections, centered title page

\documentclass{article}

\usepackage[margin=2cm]{geometry}
\usepackage{graphicx}
\usepackage{parskip}
\usepackage{url}
\usepackage{hyperref}
\usepackage{enumitem}
\usepackage{sectsty}
\usepackage{amssymb}
\usepackage{titlesec}
\usepackage{fancyhdr}
\usepackage{tikz}
\usetikzlibrary{positioning}
\usepackage{tabularx}
\newcolumntype{Y}{>{\centering\arraybackslash}X}

% === Section title formatting: LEFT-ALIGNED ===
\titleformat{\section}
  {\normalfont\Large\bfseries}{\thesection}{1em}{}

\titlespacing*{\section}{0pt}{3.5ex plus 1ex minus .2ex}{2.3ex plus .2ex}

\newcommand{\paragraphbold}[1]{\vspace{1em}\noindent\textbf{#1.}}

\pagestyle{fancy}
\fancyhf{}
\rhead{\thepage}
\lhead{COM3013 Coursework Option A Report}

\title{\bfseries Brain-Inspired Machine Unlearning: Emulating the Transformative Success of Emotional Memory Reconsolidation Therapy}
\author{
    Lam, al01759 \\[0.5em]
    \small School of Computer Science and Electronic Engineering \\
    \small University of Surrey \\
    \small Guildford, United Kingdom
}
\date{1 Feb 2025}

\begin{document}

\maketitle
\thispagestyle{empty}
\clearpage{}
\setcounter{page}{1}

\section{Abstract}
Reconsolidation is a replicable process in neurobiology to treat memory recall that is emotionally disruptive. 
Inspired by a distributive network able to cleanly target a memory and modify it, we investigate the practical models
of associative memory and attempt to create a machine unlearning model, based on it's properties.

\section{Introduction}
Machine unlearning concerns the removal of data trained into a model, with rich regulatory and law implications. 
The most notable refers to the UK General Data Protection Regulation Article 17 (the right to forget), 
protecting data ownership, consent, personal safety among others~\cite{edps2025,liu2024}. 
This task is incredibly daunting due to the entangled sequential transformation during training, 
akin to a one-way cryptographic function~\cite{cao2015}. To date, computationally acceptable solutions are mere 
approximations with degrees of residual influence, while exact solutions are far from feasibly operational~\cite{xu2023,wang2024}.

Memory reconsolidation research might provide a significant piece to the puzzle with their profound breakthrough on emotional memory 
therapy, breaking the long standing understanding that emotional memory consolidation is permanent through synaptic 
lock~\cite{brunet2008,brunet2014,brunet2018,wright2021}. Most significant is how clean and permanent this form of therapy effected memory~\cite{kamboj2020}.
(todo: what you have done)

Emulating this in context of a learnable model would have profound implications. 
Firstly, the recall and memory requirement for re-consolidation would suggest that a data address relationship exists in such models~\cite{ramsauer2020}. 
Secondly, reconsolidation allows for memory alteration, allowing data dynamic models. And thirdly, this data-targeted approach is likely computationally feasible.
iiid

\section{Literature Review}
\subsubsection{machine unlearning}
\subsubsection{current machine unlearning models and techniques}
The rapid demand for machine unlearning has led several efforts to survey and unify the field under common taxonomies, and numerous techniques for certain contexts has been developed.

The survey, "A survey of machine unlearning in large language models: Methods, challenges and future directions"
Proposed a new taxamony in classifying unlearning techniques, evaluation measures, benchmarking and benchmarking. And explores the effects of multi-modal unlearning.
Machine unlearning methods are classified into: Direct Fine-Tuning, Localized Parameter Modification, Leveraging auxiliary models, and Input/Output-based unlearning.
Evaluation measures are classified into: Adversarial Evaluation and Traditional evaluation. It also discuesses Unimodal and Multimodal benchmarking.

The servey, "Machine Unlearning: A comprehensive servey" proides a high level classification based on 4 appoaches to machine unlearning: Centralized unlearning, Federated unlearning, 
unlearning verification, and privacy and serucity issues in machine unlearning. It seperates centralized unlearning into exact unlearning, and approximate unlearning.
Below is a high level summary of current approaches in unlearning, verification and evaluation, and benchmarking.

\paragraphbold{Exact unlearning}  These methods aim to remove the influence of the forget set as if the model has been completely retrained.
They limit data influence by partitioning the dataset into shards, trained into sub-models, forming an ensemble. Unlearning is achieved by retraining,
only the affected sub-models, reducing computational cost. Although exact, these methods are still impractical in scaling, accuracy degredation and design complexity.
SISA, ARCANE and others follow this methodology.

\paragraphbold{Approximate unlearning}
Approximate unlearning is a compromise for computational efficiency with a more inherent risk of residual influence. They only provide weak unlearning guarantees.

\paragraphbold{Direct Fine-tuning}
Direct fine-tuning methods perform additional training steps on the forget set, like fine-tuning. These methods are computationally simple, but is prone to cause unintentional damage existing knowledge,
akin to catastrophic forgetting in continual learning. Below are some examples:

Gradient ascent (GA), additionally train a loss over the forget set using negative log-likelihood loss, 
To maintain model accuracy, regularization loss over the retrain set is often used, like a gradient descent or KL divergence. 

Reinforcement learning (RL) methods like Proximal Policy Optimization (PPO), Preference optimization (PO), Direct Preference optimization (DPO), negative preference optimization (NPO) that treats the forget set as preferences to policy on the model.

\paragraphbold{Localized Parameter Modification}
Localized Parameter Modification targets and adjusts specific parts of the model responsible for the forget set. These methods are precise, but require a lot of assumptions on how knowledge is located in the model. Below are some examples:
Representation engineering, like RMU, which redirects low layer forget set token to noise vector.
Locate-then-unlearn methods

\paragraphbold{Leveraging auxiliary models}
Leveraging auxiliary models typically involve using additional models to help remove or suppress forgetset influence.

\paragraphbold{Input/Output-based unlearning}
Input/Output-based unlearning focuses filtering the model's inputs or outputs without changing the core model to achieve unlearning on the forget set.


TODO: only problems and directions in machine unlearning, 
TODO: and machine unlearning verification methods, which one of that would you choose for your shit


\subsubsection{metric on machine unlearning}
NeuralIPS 2023 unlearning competition benchmark

\subsubsection{reconsolidation and emotional memory therapy}
reconsolidation (neurobiology), connect properties observed and attempts to model it.

\subsubsection{multiple-trace theory and predictive coding}
multiple-trace theory (neurobiology)
predictive coding (machine learning)

\subsubsection{associative memory modeling and unlearning}
associative memory modeling (machine learning)
unlearning (machine learning)

\subsection{key problems and challenges}
narrow down to specific problem,

\subsection{past work and state-of-the-art}
what other work has been done on it or aspects relating to,
key technologies and state-of-the-art

\section{System requirements and specification}
requirements gathering and prioritisation

appropriate software development life cycle methodologies for development projects

identify key and optional features / aspects

\section{Legal, Social Ethical and Professional issues}

\section{System Design}
apply chosen methodology (development)

design experiemnts and analytical techniques (research)

identify phases of development or experimental work

add details of design or experiments - challenges, problems, workarounds, adaptations, extensions, etc.

justify all design choices and highlight the challenges

\section{Experimental Results / Evaluation}

evaluate how the system performs

discuss results and show insights into why you got that result

\section{Conclusion and Future Work}

Overall project conclusions (how did you achieve the project objectives set at the beginning?)

Should be readable almost in isolation (reader should get a good understanding of what was achieved)

Future work - any problems that could be fixed, or scope for improvement (development), or interesting ways forward (research)

\section{References}

\begin{thebibliography}{26}

\bibitem{openai2024}
OpenAI, ``GPT-4o System Card,'' OpenAI, Tech. Rep., Sep. 2024. [Online]. Available: https://openai.com/index/gpt-4o-system-card/

\bibitem{anthropic2024}
Anthropic, ``Claude 3.5 Technical Report,'' Anthropic, Tech. Rep., Oct. 2024. [Online]. Available: https://www.anthropic.com/news/claude-3-5-sonnet

\bibitem{meta2024}
Meta AI, ``Llama 3.1-405B Release Notes and System Card,'' Meta AI, Tech. Rep., Jul. 2024. [Online]. Available: https://ai.meta.com/blog/meta-llama-3-1/

\bibitem{deepmind2024}
Google DeepMind, ``Gemini 1.5 Pro Technical Report,'' Google DeepMind, Tech. Rep., Aug. 2024. [Online]. Available: https://deepmind.google/technologies/gemini/report/

\bibitem{touvron2023}
A.~Touvron \emph{et al.}, ``Llama 2: Open foundation and fine-tuned chat models,'' Meta AI, arXiv:2307.09288v3, Jul. 2023 (updated 2024). [Online]. Available: https://arxiv.org/abs/2307.09288

\bibitem{bourtoule2021}
C.~Bourtoule \emph{et al.}, ``Machine unlearning,'' in \emph{Proc. IEEE Symp. Security Privacy (SP)}, San Francisco, CA, USA, 2021, pp. 141--159, doi: 10.1109/SP40001.2021.00092.

\bibitem{kurmanji2024}
M.~Kurmanji \emph{et al.}, ``Limits of gradient-based unlearning in large models,'' in \emph{Proc. NeurIPS}, New Orleans, LA, USA, 2024.

\bibitem{shao2025}
R.~Shao \emph{et al.}, ``Verifiable unlearning with zero-knowledge proofs,'' in \emph{Proc. ICLR}, Vienna, Austria, 2025.

\bibitem{liu2024}
K.~Z.~Liu, ``Machine unlearning in 2024,'' Stanford CS Blog, May 2024. [Online]. Available: https://ai.stanford.edu/~kzliu/blog/unlearning

\bibitem{edps2025}
European Data Protection Supervisor (EDPS), ``Preliminary Findings on Machine Unlearning and GDPR Article 17 Enforcement Actions (2024--2025),'' EDPS, Brussels, Belgium, Tech. Rep., Mar. 2025.

\bibitem{cao2015}
Y.~Cao and J.~Yang, ``Towards making systems forget with machine unlearning,'' in \emph{Proc. IEEE Symp. Security Privacy (S\&P)}, San Jose, CA, USA, May 2015, pp. 463--480, doi: 10.1109/SP.2015.35.

\bibitem{sekhar2021}
A.~Sekhari, J.~Acharya, G.~Kamath, and A.~T.~Suresh, ``Remember what you want to forget: Algorithms for machine unlearning,'' in \emph{Adv. Neural Inf. Process. Syst. (NeurIPS)}, vol. 34, 2021, pp. 18075--18086.

\bibitem{xu2023}
H.~Xu, T.~Zhu, L.~Zhang, W.~Zhou, and P.~S.~Yu, ``Machine unlearning: A survey,'' \emph{arXiv preprint arXiv:2306.03558}, 2023.

\bibitem{wang2024}
C.~Wang \emph{et al.}, ``Machine unlearning: A comprehensive survey,'' \emph{arXiv preprint arXiv:2405.07406}, 2024.

\bibitem{brunet2008}
A.~Brunet, S.~P.~Orr, J.~Tremblay, K.~Robertson, K.~Nader, and R.~K.~Pitman, ``Effect of post-retrieval propranolol on psychophysiologic responding during subsequent script-driven traumatic imagery in post-traumatic stress disorder,'' \emph{J. Psychiatr. Res.}, vol. 42, no. 6, pp. 503--506, May 2008, doi: 10.1016/j.jpsychires.2007.05.006.

\bibitem{brunet2014}
A.~Brunet, É.~Thomas, D.~Saumier, A.~R.~Ashbaugh, and R.~K.~Pitman, ``Trauma reactivation plus propranolol is associated with durably low physiological responding during subsequent script-driven traumatic imagery,'' \emph{Can. J. Psychiatry}, vol. 59, no. 4, pp. 228--232, Apr. 2014, doi: 10.1177/070674371405900408.

\bibitem{brunet2018}
A.~Brunet \emph{et al.}, ``Reduction of PTSD symptoms with pre-reactivation propranolol therapy: A randomized controlled trial,'' \emph{Am. J. Psychiatry}, vol. 175, no. 5, pp. 427--433, May 2018, doi: 10.1176/appi.ajp.2017.17050481.

\bibitem{wright2021}
L.~A.~Wright, L.~Horstmann, E.~A.~Holmes, and J.~I.~Bisson, ``Consolidation/reconsolidation therapies for the prevention and treatment of PTSD and re-experiencing: a systematic review and meta-analysis,'' \emph{Transl. Psychiatry}, vol. 11, no. 1, p. 453, 2021, doi: 10.1038/s41398-021-01570-w.

\bibitem{kamboj2020}
S.~K.~Kamboj \emph{et al.}, ``Reduction in the occurrence of distressing involuntary memories following propranolol or hydrocortisone in healthy women,'' \emph{Psychol. Med.}, vol. 50, pp. 1148--1155, 2020, doi: 10.1017/S0033291719000947.

\bibitem{smith2005}
K.~T.~Smith and L.~R.~Squire, ``Effects of memory consolidation on human hippocampal activity during retrieval,'' \emph{J. Neurosci.}, vol. 25, no. 47, pp. 10805--10810, Nov. 2005, doi: 10.1523/JNEUROSCI.2891-05.2005.

\bibitem{ngo2014}
H.-G.~Ngo, A.-M.~Björk, and J.~Born, ``Memory consolidation by replay of stimulus-specific neural activity,'' \emph{Proc. Natl. Acad. Sci. U.S.A.}, vol. 111, no. 52, pp. 18727--18732, Dec. 2014, doi: 10.1073/pnas.1418931111.

\bibitem{davis2010}
R.~Davis, S.~Renaudineau, R.~Poirier, B.~Poucet, E.~Save, and S.~Laroche, ``The formation and stability of recognition memory: What happens upon recall?,'' \emph{Front. Behav. Neurosci.}, vol. 4, p. 117, 2010, doi: 10.3389/fnbeh.2010.00117.

\bibitem{spens2023}
E.~Spens and N.~Burgess, ``A generative model of memory construction and consolidation,'' \emph{Nat. Hum. Behav.}, vol. 7, pp. 1035--1046, 2023, doi: 10.1038/s41562-023-01599-8.

\bibitem{sun2023}
W.~Sun, M.~Advani, N.~Spruston, A.~Saxe, and J.~E.~Fitzgerald, ``Organizing memories for generalization in complementary learning systems,'' \emph{Nat. Neurosci.}, vol. 26, pp. 1436--1447, 2023, doi: 10.1038/s41593-023-01381-8.

\bibitem{mcgaugh2000}
J.~L.~McGaugh, ``Memory--a century of consolidation,'' \emph{Science}, vol. 287, no. 5451, pp. 248--251, Jan. 2000, doi: 10.1126/science.287.5451.248.

\bibitem{neel2020}
S.~Neel, A.~Roth, and S.~Sharifi-Malvajerdi, ``Descent-to-delete: Gradient-based methods for machine unlearning,'' \emph{arXiv preprint arXiv:2007.02923}, 2020.

\bibitem{meng2022}
K.~Meng, D.~Bau, A.~Andonian, and Y.~Belinkov, ``Locating and editing factual associations in GPT,'' in \emph{Adv. Neural Inf. Process. Syst. (NeurIPS)}, 2022.

\bibitem{meng2023}
K.~Meng, A.~S.~Sharma, A.~Andonian, Y.~Belinkov, and D.~Bau, ``MEMIT: Mass-editing memory in a transformer,'' in \emph{Int. Conf. Learn. Represent. (ICLR)}, 2023.

\bibitem{liu2020mask}
Y.~Liu \emph{et al.}, ``Learn to forget: Machine unlearning via neuron masking,'' \emph{arXiv preprint arXiv:2003.10933}, 2020.

\bibitem{liu2023fisher}
Y.~Liu, C.~Sun, Y.~Wu, and A.~Zhou, ``Unlearning with Fisher masking,'' \emph{arXiv preprint arXiv:2310.05331}, 2023.

\bibitem{hu2021}
E.~J.~Hu, Y.~Shen, P.~Wallis, Z.~Allen-Zhu, Y.~Li, S.~Wang, L.~Wang, and W.~Chen, ``LoRA: Low-rank adaptation of large language models,'' \emph{arXiv preprint arXiv:2106.09685}, 2021.

\bibitem{yao2023}
Y.~Yao, ``Large language model unlearning,'' \emph{arXiv preprint arXiv:2310.10683}, 2023.

\bibitem{gundavarapu2024}
S.~K.~Gundavarapu, S.~Agarwal, A.~Arora, and C.~T.~Jagadeeshaiah, ``Machine unlearning in large language models,'' \emph{arXiv preprint arXiv:2405.15152}, 2024.

\bibitem{ramsauer2020}
H.~Ramsauer \emph{et al.}, ``Hopfield networks is all you need,'' \emph{arXiv preprint arXiv:2008.02217}, 2020.

\end{thebibliography}

\appendix
\section{Appendix}

\end{document}